Capabilities, MMIO and Robust Safety

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO notes: better description of how we handle Worlds + remarks

\DeclareMathOperator{\initOKo}{init_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}

\newcommand{\app}{\mathbin{+\!\!+}}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
       \\
     \\
    (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

@@

\begin{remark}
  We could simplify the theorem statement by only considering the case of a
  single entrypoint. This is equivalent: a piece of code that wants to expose
  two distinct ``methods'' can implement a single toplevel entrypoint that
  immediately returns two pointers that can then be used to invoke the two
  methods.
  %
  This would make the toplevel theorem simpler, at the cost of somewhat more
  contrieved calling convention between trusted and untrusted code.
\end{remark}

@@

We move to a theorem statement that correctly constrains the adversary in two
steps:
%
\begin{enumerate}
  \item We add a syntactic condition (on the state of memory and registers) to
    the tentative statement of our robust safety theorem, ensuring that the
    adversary code cannot gain access to any non-$\X{E}$ capabilities pointing into
    the driver code or any capabilities pointing into $\MMIO$ directly.
%
  \item Then, we provide an end-to-end statement that assumes the existence of a
    piece of boot code, that gets to run first when the capability machine
    starts up, and makes sure the above condition indeed holds before passing
    control to the adversary.

@@

\newcommand{\notMapsToR}[2]{\rightarrow^{\overline{#1}}\!\!(#2)}

Given this definition, we can now formalize the initialization constraints on
registers $r$ and memory $m$, given the region $\MMIO$ and
the driver's address space $[b,e)$ as follows:

@@

and subsequently formulate the robust safety theorem as follows:

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
         \\
   \initOK{[b,e)} \\
  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}

@@

Notice how the syntactic check for the registers is only performed on the set of registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to the driver address space.

@@

\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}

@@

It should be noted that a more general, semantic version of this specification
is also possible. It would contain the following elements, that should all
follow directly from the specification above:
\begin{itemize}
\item All registers except \X{PC} are in the value relation with respect to
  either the empty world (if we do not want to allow any non-empty, non-enter
  capabilities in registers), or a world containing a single standard region for
  all of the adversary's memory.
  The \X{PC} register satisfies both the read- and write conditions (note the
  parallel with the fundamental theorem) with respect to this same world, if we
  picked the non-empty world, or just points to the adversary's memory, which
  does not contain capabilities (if we
  picked the empty world, a capability pointing into memory can never be valid).
  It feels slightly less artificial to pick the non-empty world option here.
\item We have possession of the predicates $\I{region}$ and
  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
  need these to apply the fundamental theorem.
\item A description of the exact driver code, still in terms of points-to chunks
  (or, in this case equivalently, an invariant)
\end{itemize}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{P(t)}}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
         \\
   \X{bootSpec}(P,g,b,e,A_{\X{entry}}) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

@@

\begin{remark}
  On the level of the operational semantics, proving the boot specification
  corresponds to proving that the following holds, in order to use our concrete
  driver implementation with the robust safety theorem above (the driver still
  needs to be proven safe separately):
\[
  \begin{array}{l}
    (\textsc {BootCodeOK}) \\[0.5em]
%
    \exists n\; b\; e\ldotp (r_0, m_0, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
    := (\X{E}, g, b, e, a_i)}], m', t) \\[0.5em]
    {} \land P(t) \land \initOK{[b,e)}[r'][m']
  \end{array}
\]
\end{remark}

\subsubsection{Concrete boot and driver code}
\label{subsec:driverimpl}
Now we have a closer look at the concrete example we would like to verify,
including what its boot code looks like. Our goal is to prove that this concrete
example satisfies \X{bootSpec} (alternatively, \textsc{BootCodeOk}) and this
last robust safety theorem.

>>>>

\DeclareMathOperator{\initOKo}{init_{OK}}
\DeclareMathOperator{\initOKos}{init^{\X{sem}}_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\newcommandx{\initOKS}[4][2=r,3=m,4=W]{\initOKos(#1,#2,#3,#4)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\DeclareMathOperator{\initBCos}{init^{\X{sem}}_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
\newcommandx{\initBCS}[3][1=r,2=m,3=W]{\initBCos(#1,#2,#3)}
\DeclareMathOperator{\driverC}{code}

\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
\newcommand{\app}{\mathbin{+\!\!+}}

@@

\newcommand{\Wpub}{\sqsupseteq^{\X{pub}}}
\newcommand{\Wpriv}{\sqsupseteq^{\X{priv}}}
\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
       \\
     \\
    (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

@@

\begin{remark}
  We could simplify the theorem statement by only considering the case of a
  single entrypoint. This is equivalent: a piece of code that wants to expose
  two distinct ``methods'' can implement a single toplevel entrypoint that
  immediately returns two pointers that can then be used to invoke the two
  methods.
  %
  This would make the toplevel theorem simpler, at the cost of somewhat more
  contrieved calling convention between trusted and untrusted code.
  Another alternative is to provide an opcode to specify the operation that
  needs to be performed.
  This is slightly less general, since it does not allow providing read access
  only to an adversary.
\end{remark}

@@

We move to a theorem statement that correctly constrains the adversary in two
steps:
%
\begin{enumerate}
  \item We add a syntactic condition (on the state of memory and registers) to
    the tentative statement of our robust safety theorem, ensuring that the
    adversary code cannot gain access to any non-$\X{E}$ capabilities pointing into
    the driver code or any capabilities pointing into $\MMIO$ directly.
    Alternatively, we can formulate this condition in a semantic way, see below.
    Additionally, we specify what exact code the driver contains, so that we can
    use this in our proof that the driver is secure. This will involve adding an
    invariant to the robust safety theorem.
%
  \item Then, we provide an end-to-end statement that assumes the existence of a
    piece of boot code, that gets to run first when the capability machine
    starts up, and makes sure the above condition indeed holds before passing
    control to the adversary.

@@

\newcommand{\notMapsToR}[2]{\rightarrow^{\overline{#1}}\!\!(#2)}
\newcommand{\Winit}{\X{W_{init}}}
\newcommand{\Wemp}{\emptyset}

Given this definition, we can now formalize the initialization constraints on
registers $r$ and memory $m$, given the region $\MMIO$ and
the driver's address space $[b,e)$ as follows (for a resource-aware formulation;
see the bootSpec postcondition below):

@@

It should be noted that a more general, semantic version of this specification
is also possible. It would contain the following elements, that should all
follow directly from the specification above (TODO: write this out properly and
do the same for the semantic bootspec):
\begin{itemize}
\item All registers except \X{PC} are in the value relation with respect to
  either the empty world $\Wemp$ (if we do not want to allow any non-empty, non-enter
  capabilities in registers), or a world $\Winit$ containing a single standard region for
  all of the adversary's memory.
  The \X{PC} register satisfies both the read- and write conditions (note the
  parallel with the fundamental theorem) with respect to this same world, if we
  picked the non-empty world, or just points to the adversary's memory, which
  does not contain capabilities (if we
  picked the empty world, a capability pointing into memory can never be valid).
  It feels slightly less artificial to pick the non-empty world option here.
\item We have possession of the predicates $\I{region}$ and
  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
  need these to apply the fundamental theorem.
\item A description of the exact driver code, still in terms of points-to chunks
  (or, in this case equivalently, an invariant). We can write this invariant as
      $\driverC(b,e,\overline{\V{op}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{op}_l}}}$, with $\V{op}_l$
      some hard-coded driver instruction at location $l$.
\end{itemize}

and subsequently formulate the robust safety theorem as follows (using the
semantic specification for $\initOKo$, which references the initial world):

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
    \driverC(b,e,\overline{\V{op}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
        \end{array}} \right. \\
   \initOKS{[b,e)}[r][m][\Winit] \\
  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}

@@

Notice how the syntactic check for the registers is only performed on the set of
registers $r$, allowing the closures in $\overline{r_{a_i}}$ to point into to
the driver address space (TODO: we should change this though and add a
requirement to $\initOKo$).
(TODO: introduce notation for trace-owndership and predicate P holding)

@@

@@

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  {  \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
    \driverC(b,e,\overline{\V{op}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e,A_{\X{entry}}, \Winit) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

@@

\begin{remark}
  On the level of the operational semantics, proving the boot specification
  corresponds to proving that the following holds, in order to use our concrete
  driver implementation with the robust safety theorem above (the driver still
  needs to be proven safe separately):
\[
  \begin{array}{l}
    (\textsc {BootCodeOK}) \\[0.5em]
%
    \exists n\; b\; e\ldotp (r_0, m_0, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
    := (\X{E}, g, b, e, a_i)}], m', t) \\[0.5em]
    {} \land P(t) \land \initOK{[b,e)}[r'][m']
  \end{array}
\]
\end{remark}
\begin{remark}
 Dominique made the remark that we might want to allow capabilities pointing
 into the driver code that are not enter capabilities, but can just generally be
 proven safe wrt.\ the driver code. This scenario (which we currently do not
 support out of simplicity considerations) would force us to add the driver code
 as a static region to the world. While this would work on paper, it does not in the
 current Coq formalization, since all regions that standard regions point to are
 forced to be standard as well (through the conditions in the different cases of
 the $\I{interp1}$ predicate). Essentially, this behavior is disallowed because
 the Coq development merged the notion of read- and write condition, and there
 is hence no room to define a static region which defines a stricter reading
 policy, but hence disallowes writing. More generally, it is currently
 impossible to plug non-standard regions into the world, and allow standard
 regions to depend on them in any way whatsoever. This was not a problem up
 until now, since our examples never required this sort of dependency.
\end{remark}
\subsubsection{Concrete boot and driver code}
\label{subsec:driverimpl}
Now we have a closer look at the concrete example we would like to verify,
including what its boot code looks like. Our goal is to prove that this concrete
example satisfies \X{bootSpec} (alternatively, \textsc{BootCodeOk}) and this
last robust safety theorem.
